---
# Document Setup ---------------------------------------------------------

# Set document details.
title: 'Statistical Research: Data Sources'
author: 'Rento Saijo'
date: '2026-10-06'
date-format: iso

# Configure PDF output.
format:
  pdf:
    documentclass: article
    pdf-engine: pdflatex
    fontsize: 10pt
    papersize: letter
    geometry: margin=1in
    fontfamily: lmodern
    indent: true

# Configure code output.
execute:
  echo: false
  message: false
  warning: false

# Configure LaTeX formatting.
include-in-header:
  text: |
    % Page Layout --------------------------------------------------------

    % Load packages.
    \usepackage{fancyhdr, lastpage, titling}

    % Configure paragraphs and title.
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{4pt}
    \pretitle{\begin{center}\Large\scshape}
    \preauthor{\begin{center}\large\scshape\begin{tabular}[t]{c}}

    % Configure page footer.
    \pagestyle{fancy}
    \fancyhf{}
    \fancyfoot[C]{\textsc{Page \thepage\ of \pageref*{LastPage}}}
    \renewcommand{\headrulewidth}{0pt}
---

```{r}
#| label: setup
#| include: false
# Load Bundled Inputs ----------------------------------------------------

# Read descriptive observations and provenance.
study       <- base::readRDS('../data/analysis_data.rds')
inputs      <- study$a3z$inputs
features    <- inputs$features
coverage    <- inputs$coverage
source_rows <- inputs$sourceRows
games       <- inputs$games

# Formatting Helpers -----------------------------------------------------

# Format whole-number counts.
format_count <- function(value) base::format(value, big.mark = ',', scientific = FALSE, trim = TRUE)

# Summarize Coverage -----------------------------------------------------

# Count eligible observations by position and season.
season_ids    <- base::sort(base::unique(features$seasonId))
season_labels <- base::paste0(season_ids %/% 10000L, '--', base::substr(season_ids %% 10000L, 3L, 4L))
season_table  <- base::data.frame(
  Season      = season_labels,
  Forwards    = base::vapply(season_ids, function(season) base::sum(features$seasonId == season & features$positionCode != 'D'), base::integer(1L)),
  Defensemen  = base::vapply(season_ids, function(season) base::sum(features$seasonId == season & features$positionCode == 'D'), base::integer(1L))
)
season_table$Total <- season_table$Forwards + season_table$Defensemen

# Count source-row audit outcomes.
status_levels <- base::c('Retained', 'Unresolved game', 'Unresolved player', 'Invalid event counts', 'Player exposure disagreement', 'Game exposure disagreement')
status_table  <- base::data.frame(Status = status_levels, Rows = base::vapply(status_levels, function(status) base::sum(source_rows$rowStatus == status), base::integer(1L)))
```

```{=latex}
\thispagestyle{fancy}
\vspace{-1em}\rule{\linewidth}{0.6pt}\vspace{0.5em}
```

## What one observation represents

The scoring unit is an NHL **player and regular season**. To assemble each row, we link five-on-five NHL events and shifts with the same players and games in the [All Three Zones transition workbook](https://public.tableau.com/app/profile/corey.sznajder/viz/transitionstats/Sheet1). Recorded behavior covers 2021--22 through 2024--25, and roster outcomes extend through 2025--26 for a one-year follow-up.

NHL play-by-play supplies contact and shooting events, rosters identify players, and shifts measure five-on-five exposure; Corey Sznajder's tracked records add forechecking, retrieval, and entry-defense actions. We obtain NHL records through the pinned [`nhlscraper`](https://github.com/RentoSaijo/nhlscraper) revision and use the [All Three Zones glossary](https://www.allthreezones.com/player-cardsfaq.html) for the tracked measures. Retrieved on `r inputs$provenance$collectedAt`, the workbook has SHA-256 fingerprint `r inputs$provenance$workbookSha256`.

Official NHL draft-year scouting passages describe physical traits before the included NHL seasons. Public files retain coded mentions and a [source catalog](https://github.com/RentoSaijo/CSAx_STA496/blob/main/validation/external_validation_source_catalog.csv); passage text and blinded identity keys stay outside Git.

## Matching and eligibility

We preserve the workbook's original labels while identifying games from teams, dates within three days, and NHL rosters. Names and sweater numbers help distinguish players who share a name. A game must match unambiguously and identify at least 90% of its players; we then retain one valid row per NHL player-game, collapse identical duplicates, and exclude conflicts.

Of the workbook's `r format_count(base::nrow(games))` game labels, `r format_count(base::sum(!base::is.na(games$gameId)))` resolve to `r format_count(base::length(base::unique(stats::na.omit(games$gameId))))` distinct NHL games; fifteen labels remain unresolved. The player-game audit then yields the following flow.

```{r}
#| label: tbl-source-flow
#| tbl-cap: 'Disposition of All Three Zones skater-game rows after identity, count, and exposure checks.'
#| tbl-pos: H
# Display source-row audit counts.
knitr::kable(
  base::transform(status_table, Rows = format_count(Rows)),
  format = 'latex', booktabs = TRUE, linesep = '', row.names = FALSE, align = base::c('l', 'r')
)
```

{{< pagebreak >}}

We next compare source ice time with NHL five-on-five exposure reconstructed from shift intervals. We exclude a player-game when its absolute exposure difference exceeds 120 seconds or its game's median absolute difference exceeds 60 seconds. Event counts must also be nonnegative integers, with entry denials no greater than targets. These checks retain `r format_count(base::sum(source_rows$rowStatus == 'Retained'))` player-game rows, with a median absolute exposure disagreement of `r sprintf('%.2f', stats::median(base::abs(source_rows$deltaSeconds[source_rows$rowStatus == 'Retained']), na.rm = TRUE))` seconds. We calculate rates using **NHL shift-derived minutes** and retain source minutes for audit.

At the player-season level, eligibility for scoring requires at least 300 full-season NHL minutes, listed height and weight, and a known birth date. Scoring also requires at least 150 matched five-on-five minutes, while published rankings will require 500 full-season minutes. Of `r format_count(base::nrow(coverage))` candidate player-seasons, `r format_count(base::sum(coverage$eligible))` meet the tracked-exposure threshold: `r format_count(base::sum(features$positionCode != 'D'))` forwards and `r format_count(base::sum(features$positionCode == 'D'))` defensemen. The remaining `r format_count(base::sum(!coverage$eligible))` lack enough tracked exposure.

```{r}
#| label: tbl-seasons
#| tbl-cap: 'Eligible scored player-seasons by position and behavior season.'
#| tbl-pos: H
# Display positional study coverage.
knitr::kable(
  base::transform(season_table, Forwards = format_count(Forwards), Defensemen = format_count(Defensemen), Total = format_count(Total)),
  format = 'latex', booktabs = TRUE, linesep = '', row.names = FALSE, align = base::c('l', 'r', 'r', 'r')
)
```

The included player-seasons represent `r format_count(base::length(base::unique(features$playerId)))` players. Their median matched exposure is `r sprintf('%.2f', stats::median(features$trackedMinutes[features$positionCode != 'D']))` minutes for forwards and `r sprintf('%.2f', stats::median(features$trackedMinutes[features$positionCode == 'D']))` minutes for defensemen. We model each season separately.

## Variables used to construct CSAx

Both positional models use the same six direct measures. For each, we divide a player's attributed NHL events in matched five-on-five games by that player's NHL five-on-five minutes and express the rate per 60. With positive exposure, a recorded zero remains zero.

| Measure | Recorded numerator and interpretation |
|:--|:--|
| Hits delivered | Hits attributed to the player as hitter; possession and role affect opportunities. |
| Hits received | Hits attributed to the player as recipient; exposure to contact need not mean initiating it. |
| Opponent shots blocked | Shots blocked for the defending team, excluding teammate blocks; a recorded block need not involve body contact. |
| Fights | Recorded fighting infractions, counted separately from contact penalties; events are sparse. |
| Contact penalties taken | Whitelisted contact-related infractions attributed to the player, counted as incidents rather than minutes. |
| Contact penalties drawn | Whitelisted infractions committed by opponents against the player; puck possession and skill also affect exposure. |

The contact-penalty whitelist covers boarding, charging, checking from behind, clipping, elbowing, head checks, kneeing, roughing, slew-footing, cross-checking, high-sticking, holding, holding the stick, hooking, and tripping, with recorded helmet-removal and double-minor variants. Fighting is counted separately. The exact source labels are fixed in `R/functions.R`.

Position-specific measures add information about involvement in contested space, where forwards and defensemen face different opportunities. They do not record the pressure a player faces at every moment.

{{< pagebreak >}}

| Forward measure | Unit and denominator | Interpretation |
|:--|:--|:--|
| Net-front attempt share | Located unblocked attempts at $82\leq x\leq89$ and $|y|\leq8$ feet / all located unblocked attempts | Shooting near the crease; location does not confirm pressure. |
| Tip/deflection share | Tips and deflections / shots on goal or goals with known shot type | Redirections often occur near traffic; role affects opportunities. |
| Median shot distance | Feet from goal among located unblocked attempts | Proximity of shooting involvement. |
| Backhand share | Backhands / shots on goal or goals with known shot type | Possible constrained shooting space; backhands occur elsewhere. |
| Dump-in recoveries | All Three Zones `Recoveries` per 60 matched minutes | Recovering a dump-in and making a play. |
| Forecheck pressures | All Three Zones `Forecheck Pressures` per 60 matched minutes | Pressuring a puck-carrying defender. |

| Defenseman measure | Unit and denominator | Interpretation |
|:--|:--|:--|
| Defensive-zone retrievals | All Three Zones `DZ Retrievals` per 60 matched minutes | Collecting the puck in the defensive zone and making a play. |
| Botched retrievals | All Three Zones `Botched Retrievals` per 60 matched minutes | Retrieval-related breakdowns; the tracked category can involve the receiving player. |
| Entry denial share | All Three Zones `Denials` divided by `Targets` | Preventing an opponent's attempted entry; gap control and team tactics also matter. |

The feature table retains source counts and share denominators. A share with no opportunities remains missing until training-sample imputation. The forward model does not use the undefined entry-denial share for one center-season, so all modeled primary features are present. A 20-opportunity threshold flags sparse denominators but does not guarantee reliability.

## Outcomes and interpretation

Scouting codes record earlier mentions of active physical traits such as checking and puck battles. Silence cannot establish soft play. Official NHL records provide power-play and penalty-kill shares of total regular-season ice time; recorded zeros remain distinct from missing values.

For the postseason comparison, we will count five-on-five hits delivered and received and opponent shots blocked in each team's first four playoff games. Regular-season games excluded from CSAx inputs supply the baseline; rate comparisons require positive exposure in both periods. Continuation requires at least 300 NHL minutes the next season, with minutes per game measured among returners. We will adjust for listed size, age, current games and ice time per game, five-on-five points per 60, relative shot attempts, and season.

Tracked games are selected unevenly, and player-game totals lack timestamps or linked possession sequences, so retrievals and denials remain proxies for physical involvement. We interpret them alongside scouting evidence, roles, and coverage.

## Reproduction and sources

Inputs and provenance are bundled in `data/analysis_data.rds`. Restore `renv.lock` and render this QMD with Quarto; fresh acquisition requires the [All Three Zones workbook](https://public.tableau.com/app/profile/corey.sznajder/viz/transitionstats/Sheet1) and pinned [`nhlscraper`](https://github.com/RentoSaijo/nhlscraper).
